The Talking Heads experiment

### 2015-04-29 chap12/chap12.tex
part 2 of martin haspelmath's comments

But the sceptical observer will argue that most of these experiments are about language emergence rather than 
language evolution per se. Agents are given a certain set of strategies, which thus constitutes their 
innate Language Acquisition Device, and it is then demonstrated by computer simulations or 
robotic experiments (i) how these strategies are effective to {\itshape acquire} an existing language system in a 
population, (ii) how they are effective to {\itshape extend} a language system so that it remains adaptive 
to the needs and environments of its user, and (iii) how a population can {\itshape invent and coordinate} a new language system from 
scratch by using the strategy.

@@

To clarify these issues, let us first define a distinction between emergence, language change, and evolution:\is{culture-driven language
evolution}
\begin{itemize}
\item {\itshape Language emergence} is the phenomenon whereby a language system, i.e. a set of conventions and conceptualisations
for a particular semantic domain, for example tense-aspect, colour, space or determination, arises in a population and 
becomes shared, based on a particular strategy (or a set of strategies) that is shared by all agents. 
\item {\itshape Language change} happens when the components of a language system and its conceptualisations shift, but the 
change remains within the bounds of a particular strategy. For example, the boundaries how different cases are used 
may shift (as is happening with the dative and genitive at the moment in German) or a new basic colour category and
colour word may arise and claim a region within the colour space (as happened with the word `orange' which 
started to be used as a basic colour word in 16th century Middle English to occupy a distinctive region between yellow and red). 
\item {\itshape Language evolution} occurs when new strategies arise. There is plenty of evidence from creole languages and 
from historical linguistics that this happens at a cultural time scale. Here are three examples:
\begin{enumerate}
\item Phrase structure (in the sense of 
strict word order based on abstract hierarchical patterns) emerged gradually in 
proto-Indo European, from a stage with largely free word order but a strong inflectional system \citep{VandeVelde:2011}. 
\item Word order and prepositions became the dominant strategy instead of morphological cases 
for expressing argument structure in the late Old English period \citep{VanKemenade:1987}. 
\item There is currently an evolution going on in Spanish clitics (`le', `la', `lo') whereby the etymological system 
of Standard Spanish, which uses clitics to express different cases (nominative, dative, accusative), is shifting 
to a referential system (expressing gender and number). Case differentiation is lost, but existing forms are recruited
for the new functions implied by a referential strategy  \citep{Fernandez-Ordonez:1999}. 
\end{enumerate}
\end{itemize}
It is not always obvious from the observation of language facts whether we are dealing with change (within the boundaries 
of a strategy) or with a new strategy. However, this becomes entirely clear when we try to computationally model the processes of 
acquisition, expansion or invention of the language strategy involved. We always try to understand change first using existing 
strategies until we are forced to introduce a new strategy because the differences at the conceptual or linguistic 
level are too profound to computationally handle them.

@@

Living organisms are {\itshape adaptive} to their environment and so even within 
the bounds of a given set of innate strategies there may still be significant changes observed in appearance, 
behaviour or internal functioning. For example, beavers 
may start using rocks for dams when mud and branches are not available, birds use plastics or other waste products, 
a goat may start to walk on two hind legs when it did not grow the forelegs \citep{Jablonka:2013}. 
Nevertheless true {\itshape evolution} is only said to happen if there are new innate 
strategies, i.e. genetically coded strategies, that arise, for example if a fish species genetically 
speciates into subspecies where one continues to bread eggs deposited on the bottom of the sea whereas the 
new variant does mouthbrooding, meaning that eggs are bred within the mouth of a parent. 
Also in biology it is not always clear whether we are dealing with adaptive change (without genomic change) or evolution
(with genomic change). Some variants that were thought to be a different species at one point of time
turned out to be ecotypes of the same species, i.e. variants that 
have the same genetics but a different appearance or behaviour due to adaptation to the environment. However today these debates
can be entirely settled when a genetic analysis is performed.

@@

But it is also possible that language strategies are the outcome of cognitive development and cultural evolution, 
just like humans have developed a multitude of non-innate strategies in a wide range of domains: strategies for finding your 
way in cities, for building shelters, for preparing food, for playing Jazz piano, for fixing technical equipment, 
for organising trade, composing music, negotiating a contract, finding a partner, etc.  
This is the idea behind the {\itshape recruitment theory} of language evolution, which postulates that cognition and 
cultural evolution are the primary motors for the origins and propagation of new language strategies, even though of 
course they require the necessary neurobiological capacities that make each strategy possible. Once a strategy culturally 
evolved, it exerts a pull effect on genetic evolution, because genetic variants are favoured that can carry out a strategy more easily. 
So this hypothesis implies {\itshape culture-driven gene evolution} rather than gene-driven cultural 
evolution \citep{Fisher:2013}.

@@

What could a theory of cultural evolution look like? There is certainly no 
consensus on this.\footnote{See the debate in: \cite{Richerson:2005}.}
But the most obvious choice is to view cultural evolution as another instance of a Darwinian evolutionary dynamics, happening at 
the cultural and cognitive level rather than at the genetic level. This idea has been suggested by many researchers, 
including Richard Dawkins in his theory of memes \citep{Dawkins:1976}. 
The basic principle underlying Darwinian dynamics 
is to split the process of coming up with a design in three subprocesses: 
\begin{enumerate}
\item There is a process of {\itshape generating} possible variants, often by making changes to 
existing variants or by information transfer between variants, for example through cross-over. 
\item There is a separate process of {\itshape testing} whether these variants satisfy 
desired selection criteria. This happens usually in a kind of competition where different variants 
compete for available resources. 
\item There is a {\itshape self-enforcing causal loop} between the 
outcome of selection and the frequency with which a variant is maintained and hence used to generate new 
variants. 
\end{enumerate}
The last step makes Darwinian dynamics cumulative: Partly working solutions can be built upon further and lead 
the way to even better ones.

@@

What kind of selectionist forces are relevant in the case of language? \is{selectionist forces}
Clearly, the bottom line is that language users want to achieve {\bfshape communicative success}. 
Communicative success means that the speaker achieves his or her non-linguistic goal. For example, if a speaker
wants a hearer to perform some action (such as `Can you get me the car keys?'), there is communicative success if 
the hearer indeed performs the requested action. Speakers and hearers generally do not care whether their sentences
are perfectly grammatical or complete or semantically accurate. Speakers assume that hearers are intelligent enough
to infer the missing information and that they can fill in the required details from the context. 
In normal spoken dialogue, communicative success is immediately obvious, either 
from the actions that the hearer performs as a consequence of an utterance or by gestural cues whereby 
the hearer informs the speaker that he or she is being understood.

@@

\begin{itemize}
\item {\itshape Expressive adequacy}: The language systems available to the speaker and the hearer must have 
the necessary expressive adequacy to reach communicative success. 
The available conceptual systems must include the needed conceptual distinctions 
and the available linguistic systems must be able to express these distinctions. Expressive adequacy depends 
not only on the goals of the speaker and the available embodiment but also on the environment. For example, 
if only black-and-white pictures are available, a strategy based on hue is entirely inadequate. On the other hand, 
if one of the dialogue partners is colour blind, a hue-based strategy would not work for coloured pictures. 
\is{expressive adequacy}

\item {\itshape Cognitive effort}: In order to cope with the incredible speed of normal language production and comprehension, 
reducing cognitive effort is of primordial importance. Cognitive effort is expended at all levels
of language: How much time and memory 
needs to be spent in coming up with adequate conceptualisations? How complex is the process of constructing
a sentence? How 
difficult is it to articulate the speech sounds? How efficiently can the sentence be parsed? How complicated is the 
interpretation of the sentence? If sentences are too 
complex to comprehend, for example because they trigger a large amount of combinatorial search, then 
hearers give up. Or if sentences 
are too hard to produce, the production process will be too slow to maintain the hearer's attention. 
Optimisations of cognitive effort are often in conflict for speaker and hearer. What requires less effort for the 
speaker might mean more effort for the hearer and vice versa. 
\is{cognitive effort}

\item {\itshape Learnability}: Speakers must occasionally expand their language systems to express novel meanings or 
better capture the attention of hearers with a ``fresh'' way of saying something. But these innovations will 
lead nowhere if they cannot be acquired by hearers. They will lead to communicative failure and 
the innovation will not propagate in the population and therefore it has no chance to survive in the shared language.
Learnability depends on whether speaker and hearer share the same language strategies, how much additional context 
is available, and how far the speaker has stretched established conventions to achieve his communicative goal. 
\is{learnability}

\item {\itshape Social conformity}: Speakers and hearers can greatly optimise their language production and 
comprehension by making their language systems as similar as possible. Moreover, details of language use, such 
as pronunciation, signal to which group a speaker belongs, and speakers often seek social conformity with the 
language use of their peers in order to belong to the group. Or they mimic the language systems 
of the most prestigious group in order to show they belong to that. The opposite happens as well. Speakers 
may try to mark themselves as linguistically different from a group to which they do not want to belong. 
\is{social conformity}
\end{itemize}

@@

Organisms use a variety of strategies for survival and offspring production:
bodily shape and appearance, specific sensors and actuators, metabolism, 
innate behaviours. Biologists call the ecological area that a particular organism occupies 
thanks to its physiological and behavioural strategies its ecological {\itshape niche}. A niche is defined through the ecological conditions
(temperature, altitude, habitat type, food availability, presence of predators, etc.) for which the organism's strategies
are appropriate and hence within which the species reaches a sufficiently high fitness to survive and fend off predators. 
Organisms may also have strategies to alter their environment so that some of their strategies become more possible or more optimal, 
a process called {\itshape niche construction} \citep{Odling-Smee:2003} \is{fitness landscape}

Niche analysis is performed by defining a multi-dimensional space, further called the {\itshape niche space}. Each 
dimension or {\itshape niche factor} has to be dealt with by the species in order to remain viable. There 
is consequently a direct
relation between the dimensions of the niche space and the selectionist pressures on the organisms trying to flourish 
within that niche which form a fitness landscape. Niche analysis
 typically examines how the different species within a delineated geological area 
are located within the niche space. It is then possible 
to see how species overlap, how they compete, how they are complementary, or whether some species are beginning
to occupy a niche left open by others. Usually the dimensions of the niche space are not known but have 
to be inferred through multi-variate analysis from empirical observations of species occurrence and 
detailed descriptions of environmental factors \citep{Elith:2009}.  But it is also possible to 
operate in a top-down manner, based on prior analysis of what might be niche factors, and to manipulate the environment 
to detect under what conditions the members of a species will thrive or not \citep{Nosil:2008}.

@@

Van Trijp initialised three populations of agents: with the Old High German system, 
the Middle High German system, or the New High German System. He then defined the following four dimensions for the 
fitness landscape: \enlargethispage{1\baselineskip}
\begin{enumerate}
\item {\itshape Search effort} which is the amount of search that is needed, scaled between minimum (when there is one marker for every of 
the 18 possible configurations and so no search is needed) and maximum (when there are no markers at all).
\item {\itshape Disambiguation power} which is based on the number of possible interpretations left after application of the grammar. 
The disambiguation power is also scaled with maximum power for 12 markers and minimum for 0 markers (not counting the genitive). 
\item {\itshape Articulatory effort} which is calculated in terms of the number of movements that articulators such as the lips and 
tongue have to make when pronouncing the different phonemes that make up each article. 
\item {\itshape Acoustic precision} which is based on the phonological distance between words.
\end{enumerate}
Criteria (1), (2) and (4) pertain to the hearer and (1) and (3) to the speaker. They are calculated for each utterance
in a corpus and the weighted average gives a value for the performance of the language system for each of these dimensions. The global 
performance can be assessed by taking a (weighted) average of each of the values for all dimensions. 
As expected, Old High German scores the highest for disambiguation power and acoustic precision and the lowest for articulatory effort 
and processing effort, thus being an advantage for the hearer but not for the speaker (see \figref{fig:german-system}).

@@

\begin{figure}[b]
\centerline{\includegraphics[width=0.8\linewidth]{chap11/figs/spatial.pdf}}
\caption{Set-up for the spatial language game experiments. Two robots are looking at a scene 
with a few blocks, a bigger box with tags indicating front, back or left and right side, and a global landmark in the form 
of a large tag pasted on the wall. {\itshape Top left and right:} World as seen through the camera of the robot. 
{\itshape Bottom left and right:} The internal world model as perceived by the left and right robot.}
\label{fig:spatial}
\end{figure}
\begin{enumerate}
\item The use of {\itshape projective relations} such as front/back, above/below, left/right. This requires imposing a coordinate
system on the world model and name areas within this coordinate system. A common coordinate system is to take 
an object, such as the human body of the speaker as reference point and project vectors emanating from the body. 
\item The use of {\itshape proximal relations} such as far/near. This requires that the distances of objects with respect to the 
reference point are computed and categorised. 
\item The use of {\itshape absolute relations}, such as North/West/East/South, related to global directions. 
\item The use of an {\itshape allocentric landmark} (as opposed to an egocentric one)
from which the situation is conceptualised. For example, in the 
phrase `the block in front of the box', the box is acting as a landmark and the projective spatial relations are 
computed as emanating from that object. Another example is `the block on your left' where the landmark is the 
hearer. 
\end{enumerate}

@@

Spranger used the following additional internal factors: 
\begin{enumerate}
\item {\itshape Communicative success}, including whether production failed, interpretation failed, or the hearer pointed to the wrong 
object (pragmatic failure). We see (\figref{fig:why-grammar-german}) that the more complex a situation becomes, the 
more grammar becomes relevant. 
\item {\itshape Semantic ambiguity}, being the number of semantic structures that remain after syntactic analysis and have to be mapped 
onto the perceived world in order to be disambiguated (See \figref{fig:ambiguity}, left). Without grammar there is a lot 
more semantic ambiguity and hence more cognitive effort to consult the world model. 
\item {\itshape Interpretation sharing}, which takes place when the semantic structure derived by the hearer is equal to that originally 
intended by the speaker (See \figref{fig:ambiguity}, right). Interpretation sharing is important in cases of displaced 
communication when agents do not have access to a shared world. We see that without grammar, interpretation sharing is 
heavily compromised. 
\end{enumerate}

@@

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorisation using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming Game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): \enlargethispage{2\baselineskip}
\begin{enumerate}
\item The {\itshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\itshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\itshape full-colour strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

@@

It is only one step now to achieve the selection and alignment of language strategies. Basically, two mechanisms needs to be 
added: \is{strategy selection}
\begin{enumerate}
\item Agents need to track the performance of their strategies by tracking the performance of the language 
systems that were built based on their strategies. In other words, not only lexical or grammatical constructions
but also strategies need to have a score. These {\itshape strategy scores} should be updated using the same 
lateral inhibition dynamics that we have been using at the level of word-meaning associations or 
more generally constructions. 
\item It is also necessary that the conceptualisations and the lexical and grammatical constructions that 
are implicated within a particular strategy are tagged with that strategy, so that if their score is updated, 
the score of the strategy can be updated as well (\figref{fig:tagging}). This is necessary  
for implementing a proper credit assignment 
mechanism and also for ensuring that a word becomes unambiguously used with a particular strategy. 
\end{enumerate}
The same word may be tagged with more than one strategy because 
a learning agent does not know which strategy has been used by the speaker and so this may lead to different
hypotheses. But in the end, agents strive for convergence so that they all use the 
same word with the same meaning. For example, a hearer cannot know whether a particular word (e.g. `yellow')
is to be interpreted using  the brightness-based or  the full-colour-based strategy, particularly
because both strategies could work in similar circumstances. For example, yellow colour chips are often the 
most bright ones and hence both strategies would work. Consequently it is unavoidable that different strategies compete
to recruit the same words, and this explains that we see this phenomenon indeed in the historically attested evolution 
of language. Indeed, many of today's hue words like `yellow', `brown' 
or `blue' were all expressing brightness-based distinctions in Old English before they became used as part of the  
Basic Colour Strategy in the late Middle English period (1350--1500).\footnote{See
\cite{MacLaury:1992} and \cite{Casson:1997}.}

@@

\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/full-1000.pdf} ~~~
\includegraphics[height=0.5\textwidth]{chap12/figs/full-3000.pdf}
}
\caption{{The {\itshape Basic Colour Strategy} allows a population of agents (in this case 10) to self-organise a colour lexicon from scratch. The graph shows that high communicative success is reached with a lexicon of about 15 colour words (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown after 400 (bottom left) and 1200 (bottom right) games per agent. Each row represents the lexicon of one agent.}
\label{fig:full-cs-dynamics}
} 
\end{figure}

The {\itshape Brightness Prototype Strategy} is similar to the Basic Colour Strategy, but instead of taking all three dimensions into account, only the L* dimension of both the stimuli in the context and the prototypes of the colour categories are compared. While learning, the prototype of the used colour category is shifted on the L* axis towards the L* value of the topic. During invention, only the L* value of the topic is considered relevant. \figref{fig:brightness-dynamics} shows that this strategy is also adequate to allow a population of agents to self-organise and coordinate a colour lexicon from scratch. The resulting colour lexicon now consists of different shades of grey. The set of prototypes also expands and becomes similar across agents.

@@

Rich and complex dynamics results from this general system. In the results shown in \figref{fig:strategies-one}, the 
set of prototypes is kept fixed, namely equal to the focal colours underlying 
the Spanish colour system, which can be used either for the Full-Colour strategy or for the Brightness strategy. 
For example, the word `morado' (`purple') can both be interpreted in the full-colour space and in the brightness space.

>>>>

But the sceptical observer will argue that most of these experiments are about language emergence rather than 
language evolution per se. Agents are given a certain set of strategies, which thus constitutes their 
innate Language Acquisition Device, and it is then demonstrated by computer simulations or 
robotic experiments (i) how these strategies are effective to {\scshape acquire} an existing language system in a 
population, (ii) how they are effective to {\scshape extend} a language system so that it remains adaptive 
to the needs and environments of its user, and (iii) how a population can {\scshape invent and coordinate} a new language system from 
scratch by using the strategy.

@@

To clarify these issues, let us first define a distinction between emergence, language change, and evolution:\is{culture-driven language
evolution}
\begin{itemize}
\item {\scshape Language emergence} is the phenomenon whereby a language system, i.e. a set of conventions and conceptualisations
for a particular semantic domain, for example tense-aspect, colour, space or determination, arises in a population and 
becomes shared, based on a particular strategy (or a set of strategies) that is shared by all agents. 
\item {\scshape Language change} happens when the components of a language system and its conceptualisations shift, but the 
change remains within the bounds of a particular strategy. For example, the boundaries how different cases are used 
may shift (as is happening with the dative and genitive at the moment in German) or a new basic colour category and
colour word may arise and claim a region within the colour space (as happened with the word `orange' which 
started to be used as a basic colour word in 16th century Middle English to occupy a distinctive region between yellow and red). 
\item {\scshape Language evolution} occurs when new strategies arise. There is plenty of evidence from creole languages and 
from historical linguistics that this happens at a cultural time scale. Here are three examples:
\begin{enumerate}
\item Phrase structure (in the sense of 
strict word order based on abstract hierarchical patterns) emerged gradually in 
proto-Indo European, from a stage with largely free word order but a strong inflectional system \citep{VandeVelde:2011}. 
\item Word order and prepositions became the dominant strategy instead of morphological cases 
for expressing argument structure in the late Old English period \citep{VanKemenade:1987}. 
\item There is currently an evolution going on in Spanish clitics (`le', `la', `lo') whereby the etymological system 
of Standard Spanish, which uses clitics to express different cases (nominative, dative, accusative), is shifting 
to a referential system (expressing gender and number). Case differentiation is lost, but existing forms are recruited
for the new functions implied by a referential strategy  \citep{Fernandez-Ordonez:1999}. 
\end{enumerate}
\end{itemize}
It is not always obvious from the observation of language facts whether we are dealing with change (within the boundaries 
of a strategy) or with a new strategy. However, this becomes entirely clear when we try to computationally model the processes of 
acquisition, expansion or invention of the language strategy involved. We always try to understand change first using existing 
strategies until we are forced to introduce a new strategy because the differences at the conceptual or linguistic 
level are too profound to computationally handle them.

@@

Living organisms are {\scshape adaptive} to their environment and so even within 
the bounds of a given set of innate strategies there may still be significant changes observed in appearance, 
behaviour or internal functioning. For example, beavers 
may start using rocks for dams when mud and branches are not available, birds use plastics or other waste products, 
a goat may start to walk on two hind legs when it did not grow the forelegs \citep{Jablonka:2013}. 
Nevertheless true {\scshape evolution} is only said to happen if there are new innate 
strategies, i.e. genetically coded strategies, that arise, for example if a fish species genetically 
speciates into subspecies where one continues to bread eggs deposited on the bottom of the sea whereas the 
new variant does mouthbrooding, meaning that eggs are bred within the mouth of a parent. 
Also in biology it is not always clear whether we are dealing with adaptive change (without genomic change) or evolution
(with genomic change). Some variants that were thought to be a different species at one point of time
turned out to be ecotypes of the same species, i.e. variants that 
have the same genetics but a different appearance or behaviour due to adaptation to the environment. However today these debates
can be entirely settled when a genetic analysis is performed.

@@

But it is also possible that language strategies are the outcome of cognitive development and cultural evolution, 
just like humans have developed a multitude of non-innate strategies in a wide range of domains: strategies for finding your 
way in cities, for building shelters, for preparing food, for playing Jazz piano, for fixing technical equipment, 
for organising trade, composing music, negotiating a contract, finding a partner, etc.  
This is the idea behind the {\scshape recruitment theory} of language evolution, which postulates that cognition and 
cultural evolution are the primary motors for the origins and propagation of new language strategies, even though of 
course they require the necessary neurobiological capacities that make each strategy possible. Once a strategy culturally 
evolved, it exerts a pull effect on genetic evolution, because genetic variants are favoured that can carry out a strategy more easily. 
So this hypothesis implies {\scshape culture-driven gene evolution} rather than gene-driven cultural 
evolution \citep{Fisher:2013}.

@@

What could a theory of cultural evolution look like? There is certainly no 
consensus on this.\footnote{See the debate in: \cite{Richerson:2005}.}
But the most obvious choice is to view cultural evolution as another instance of a Darwinian evolutionary dynamics, happening at 
the cultural and cognitive level rather than at the genetic level. This idea has been suggested by many researchers, 
including Richard Dawkins in his theory of memes \citep{Dawkins:1976}. 
The basic principle underlying Darwinian dynamics 
is to split the process of coming up with a design in three subprocesses: 
\begin{enumerate}
\item There is a process of {\scshape generating} possible variants, often by making changes to 
existing variants or by information transfer between variants, for example through cross-over. 
\item There is a separate process of {\scshape testing} whether these variants satisfy 
desired selection criteria. This happens usually in a kind of competition where different variants 
compete for available resources. 
\item There is a {\scshape self-enforcing causal loop} between the 
outcome of selection and the frequency with which a variant is maintained and hence used to generate new 
variants. 
\end{enumerate}
The last step makes Darwinian dynamics cumulative: Partly working solutions can be built upon further and lead 
the way to even better ones.

@@

What kind of selectionist forces are relevant in the case of language? \is{selectionist forces}
Clearly, the bottom line is that language users want to achieve {\scshape communicative success}. 
Communicative success means that the speaker achieves his or her non-linguistic goal. For example, if a speaker
wants a hearer to perform some action (such as `Can you get me the car keys?'), there is communicative success if 
the hearer indeed performs the requested action. Speakers and hearers generally do not care whether their sentences
are perfectly grammatical or complete or semantically accurate. Speakers assume that hearers are intelligent enough
to infer the missing information and that they can fill in the required details from the context. 
In normal spoken dialogue, communicative success is immediately obvious, either 
from the actions that the hearer performs as a consequence of an utterance or by gestural cues whereby 
the hearer informs the speaker that he or she is being understood.

@@

\begin{itemize}
\item {\scshape Expressive adequacy}: The language systems available to the speaker and the hearer must have 
the necessary expressive adequacy to reach communicative success. 
The available conceptual systems must include the needed conceptual distinctions 
and the available linguistic systems must be able to express these distinctions. Expressive adequacy depends 
not only on the goals of the speaker and the available embodiment but also on the environment. For example, 
if only black-and-white pictures are available, a strategy based on hue is entirely inadequate. On the other hand, 
if one of the dialogue partners is colour blind, a hue-based strategy would not work for coloured pictures. 
\is{expressive adequacy}

\item {\scshape Cognitive effort}: In order to cope with the incredible speed of normal language production and comprehension, 
reducing cognitive effort is of primordial importance. Cognitive effort is expended at all levels
of language: How much time and memory 
needs to be spent in coming up with adequate conceptualisations? How complex is the process of constructing
a sentence? How 
difficult is it to articulate the speech sounds? How efficiently can the sentence be parsed? How complicated is the 
interpretation of the sentence? If sentences are too 
complex to comprehend, for example because they trigger a large amount of combinatorial search, then 
hearers give up. Or if sentences 
are too hard to produce, the production process will be too slow to maintain the hearer's attention. 
Optimisations of cognitive effort are often in conflict for speaker and hearer. What requires less effort for the 
speaker might mean more effort for the hearer and vice versa. 
\is{cognitive effort}

\item {\scshape Learnability}: Speakers must occasionally expand their language systems to express novel meanings or 
better capture the attention of hearers with a ``fresh'' way of saying something. But these innovations will 
lead nowhere if they cannot be acquired by hearers. They will lead to communicative failure and 
the innovation will not propagate in the population and therefore it has no chance to survive in the shared language.
Learnability depends on whether speaker and hearer share the same language strategies, how much additional context 
is available, and how far the speaker has stretched established conventions to achieve his communicative goal. 
\is{learnability}

\item {\scshape Social conformity}: Speakers and hearers can greatly optimise their language production and 
comprehension by making their language systems as similar as possible. Moreover, details of language use, such 
as pronunciation, signal to which group a speaker belongs, and speakers often seek social conformity with the 
language use of their peers in order to belong to the group. Or they mimic the language systems 
of the most prestigious group in order to show they belong to that. The opposite happens as well. Speakers 
may try to mark themselves as linguistically different from a group to which they do not want to belong. 
\is{social conformity}
\end{itemize}

@@

Organisms use a variety of strategies for survival and offspring production:
bodily shape and appearance, specific sensors and actuators, metabolism, 
innate behaviours. Biologists call the ecological area that a particular organism occupies 
thanks to its physiological and behavioural strategies its ecological {\scshape niche}. A niche is defined through the ecological conditions
(temperature, altitude, habitat type, food availability, presence of predators, etc.) for which the organism's strategies
are appropriate and hence within which the species reaches a sufficiently high fitness to survive and fend off predators. 
Organisms may also have strategies to alter their environment so that some of their strategies become more possible or more optimal, 
a process called {\scshape niche construction} \citep{Odling-Smee:2003} \is{fitness landscape}

Niche analysis is performed by defining a multi-dimensional space, further called the {\scshape niche space}. Each 
dimension or {\scshape niche factor} has to be dealt with by the species in order to remain viable. There 
is consequently a direct
relation between the dimensions of the niche space and the selectionist pressures on the organisms trying to flourish 
within that niche which form a fitness landscape. Niche analysis
 typically examines how the different species within a delineated geological area 
are located within the niche space. It is then possible 
to see how species overlap, how they compete, how they are complementary, or whether some species are beginning
to occupy a niche left open by others. Usually the dimensions of the niche space are not known but have 
to be inferred through multi-variate analysis from empirical observations of species occurrence and 
detailed descriptions of environmental factors \citep{Elith:2009}.  But it is also possible to 
operate in a top-down manner, based on prior analysis of what might be niche factors, and to manipulate the environment 
to detect under what conditions the members of a species will thrive or not \citep{Nosil:2008}.

@@

Van Trijp initialised three populations of agents: with the Old High German system, 
the Middle High German system, or the New High German System. He then defined the following four dimensions for the 
fitness landscape: \enlargethispage{1\baselineskip}
\begin{enumerate}
\item {\scshape Search effort}, which is the amount of search that is needed, scaled between minimum (when there is one marker for every of 
the 18 possible configurations and so no search is needed) and maximum (when there are no markers at all).
\item {\scshape Disambiguation power}, which is based on the number of possible interpretations left after application of the grammar. 
The disambiguation power is also scaled with maximum power for 12 markers and minimum for 0 markers (not counting the genitive). 
\item {\scshape Articulatory effort}, which is calculated in terms of the number of movements that articulators such as the lips and 
tongue have to make when pronouncing the different phonemes that make up each article. 
\item {\scshape Acoustic precision}, which is based on the phonological distance between words.
\end{enumerate}
Criteria (1), (2) and (4) pertain to the hearer and (1) and (3) to the speaker. They are calculated for each utterance
in a corpus and the weighted average gives a value for the performance of the language system for each of these dimensions. The global 
performance can be assessed by taking a (weighted) average of each of the values for all dimensions. 
As expected, Old High German scores the highest for disambiguation power and acoustic precision and the lowest for articulatory effort 
and processing effort, thus being an advantage for the hearer but not for the speaker (see \figref{fig:german-system}).

@@

\begin{figure}[b]
\centerline{\includegraphics[width=0.8\linewidth]{chap11/figs/spatial.pdf}}
\caption{Set-up for the spatial language game experiments. Two robots are looking at a scene 
with a few blocks, a bigger box with tags indicating front, back or left and right side, and a global landmark in the form 
of a large tag pasted on the wall. {\itshape Top left and right:} World as seen through the camera of the robot. 
{\itshape Bottom left and right:} The internal world model as perceived by the left and right robot.}
\label{fig:spatial}
\end{figure}
\begin{enumerate}
\item The use of {\scshape projective relations} such as front/back, above/below, left/right. This requires imposing a coordinate
system on the world model and name areas within this coordinate system. A common coordinate system is to take 
an object, such as the human body of the speaker as reference point and project vectors emanating from the body. 
\item The use of {\scshape proximal relations} such as far/near. This requires that the distances of objects with respect to the 
reference point are computed and categorised. 
\item The use of {\scshape absolute relations}, such as North/West/East/South, related to global directions. 
\item The use of an {\scshape allocentric landmark} (as opposed to an egocentric one)
from which the situation is conceptualised. For example, in the 
phrase `the block in front of the box', the box is acting as a landmark and the projective spatial relations are 
computed as emanating from that object. Another example is `the block on your left' where the landmark is the 
hearer. 
\end{enumerate}

@@

Spranger used the following additional internal factors: 
\begin{enumerate}
\item {\scshape Communicative success}, including whether production failed, interpretation failed, or the hearer pointed to the wrong 
object (pragmatic failure). We see (\figref{fig:why-grammar-german}) that the more complex a situation becomes, the 
more grammar becomes relevant. 
\item {\scshape Semantic ambiguity}, being the number of semantic structures that remain after syntactic analysis and have to be mapped 
onto the perceived world in order to be disambiguated (See \figref{fig:ambiguity}, left). Without grammar there is a lot 
more semantic ambiguity and hence more cognitive effort to consult the world model. 
\item {\scshape Interpretation sharing}, which takes place when the semantic structure derived by the hearer is equal to that originally 
intended by the speaker (See \figref{fig:ambiguity}, right). Interpretation sharing is important in cases of displaced 
communication when agents do not have access to a shared world. We see that without grammar, interpretation sharing is 
heavily compromised. 
\end{enumerate}

@@

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorisation using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming Game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): \enlargethispage{2\baselineskip}
\begin{enumerate}
\item The {\scshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\scshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\scshape full-colour strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

@@

It is only one step now to achieve the selection and alignment of language strategies. Basically, two mechanisms needs to be 
added: \is{strategy selection}
\begin{enumerate}
\item Agents need to track the performance of their strategies by tracking the performance of the language 
systems that were built based on their strategies. In other words, not only lexical or grammatical constructions
but also strategies need to have a score. These {\scshape strategy scores} should be updated using the same 
lateral inhibition dynamics that we have been using at the level of word-meaning associations or 
more generally constructions. 
\item It is also necessary that the conceptualisations and the lexical and grammatical constructions that 
are implicated within a particular strategy are tagged with that strategy, so that if their score is updated, 
the score of the strategy can be updated as well (\figref{fig:tagging}). This is necessary  
for implementing a proper credit assignment 
mechanism and also for ensuring that a word becomes unambiguously used with a particular strategy. 
\end{enumerate}
The same word may be tagged with more than one strategy because 
a learning agent does not know which strategy has been used by the speaker and so this may lead to different
hypotheses. But in the end, agents strive for convergence so that they all use the 
same word with the same meaning. For example, a hearer cannot know whether a particular word (e.g. `yellow')
is to be interpreted using  the brightness-based or  the full-colour-based strategy, particularly
because both strategies could work in similar circumstances. For example, yellow colour chips are often the 
most bright ones and hence both strategies would work. Consequently it is unavoidable that different strategies compete
to recruit the same words, and this explains that we see this phenomenon indeed in the historically attested evolution 
of language. Indeed, many of today's hue words like `yellow', `brown' 
or `blue' were all expressing brightness-based distinctions in Old English before they became used as part of the  
Basic Colour Strategy in the late Middle English period (1350--1500).\footnote{See
\cite{MacLaury:1992} and \cite{Casson:1997}.}

@@

\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/full-1000.pdf} ~~~
\includegraphics[height=0.5\textwidth]{chap12/figs/full-3000.pdf}
}
\caption{{The {\scshape Basic Colour Strategy} allows a population of agents (in this case 10) to self-organise a colour lexicon from scratch. The graph shows that high communicative success is reached with a lexicon of about 15 colour words (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown after 400 (bottom left) and 1200 (bottom right) games per agent. Each row represents the lexicon of one agent.}
\label{fig:full-cs-dynamics}
} 
\end{figure}

The {\scshape Brightness Prototype Strategy} is similar to the Basic Colour Strategy, but instead of taking all three dimensions into account, only the L* dimension of both the stimuli in the context and the prototypes of the colour categories are compared. While learning, the prototype of the used colour category is shifted on the L* axis towards the L* value of the topic. During invention, only the L* value of the topic is considered relevant. \figref{fig:brightness-dynamics} shows that this strategy is also adequate to allow a population of agents to self-organise and coordinate a colour lexicon from scratch. The resulting colour lexicon now consists of different shades of grey. The set of prototypes also expands and becomes similar across agents.

@@

Rich and complex dynamics results from this general system. In the results shown in \figref{fig:strategies-one}, the 
set of prototypes is kept fixed, namely equal to the focal colours underlying 
the Spanish colour system, which can be used either for the Full-Colour strategy or for the Brightness strategy. 
For example, the word \textit{morado} (`purple') can both be interpreted in the full-colour space and in the brightness space.
